Analysis-ready VCF at Biobank scale using Zarr

### 2024-04-30 paper.tex
Some more Zarr refs

% tentative
\title{Columnar VCF for Biobank scale datasets using Zarr}

@@

\subsection{Zarr and block-based compression}
% Zarr is a simple layer on top of best-in-class, modern components
In the interest of completeness it is useful to provide a high-level overview
of Zarr and the techologies that it depends upon. Zarr is a specialised format
for storing large-scale $n$-dimensional numerical data (arrays). Arrays
are split into chunks, which are compressed and stored separately. Chunks are 
addressed by their indexes along the dimensions of the array, and the 
compressed data associated with this key. Chunks are
are often stored in individual files, but a wide array of different
stores are supported including cloud buckets and NoSQL databases. 
Metadata describing the array and its properties is then stored 
in JSON format along with the chunks. Zarr is best seen as a 
\emph{specification} that describes the details of how chunks are 
addressed and stored: there are multiple implementations across
different languages.

@@

@@

>>>>

% See https://github.com/sgkit-dev/vcf-zarr-publication/issues/87
% for discussion
\title{Analysis-ready VCF at Biobank scale using Zarr}

@@

\subsection{Zarr and block-based compression}
% Zarr is a simple layer on top of best-in-class, modern components
In the interest of completeness it is useful to provide a high-level overview
of Zarr and the techologies that it depends upon. Zarr is a specialised format
for storing large-scale $n$-dimensional numerical data (arrays). Arrays
are split into chunks, which are compressed and stored separately. Chunks are 
addressed by their indexes along the dimensions of the array, and the 
compressed data associated with this key. Chunks are
are often stored in individual files, but a wide array of different
stores are supported including cloud buckets and NoSQL databases. 
In principle, Zarr can store data into any key-value store.
Metadata describing the array and its properties is then stored 
in JSON format along with the chunks. Zarr is best seen as a 
\emph{specification} that describes the details of how chunks are 
addressed and stored: there are multiple implementations across
different languages.

% Good points from  the new Horizons technical paper:

% From a technical standpoint, a standout feature of Zarr is its simplicity.
% The format is easy for developers to
% understand because the internals are exposed transparently. Consequently, Zarr
% is considered easy to “hack,”
% i.e. to adapt to new use cases and scenarios, such as implementing a reader
% from a new programming language
% or creating a new backend storage layer. (This is a significant contrast from
% NetCDF4 / HDF5, whose internal file
% format is relatively opaque, and for which all access must go through the
% official C library.)

% The Zarr project is governed by a steering council of contributors from
% different institutions. No single company or institution controls the
% specification and its implementations.
% Second, although Zarr is relatively new, it is currently being evaluated by the
% Open Geospatial Consortium (

The Open GeoSpatial consortium has formally recognised Zarr as a community
standard~\cite{ogc_zarr2_standard}
and has formed
a new GeoZarr Standards Working Group to establish a Zarr encoding for
geospatial data~\cite{ogc_geozarr_news}.

Zarr is being considered a potential format for migrating the 
Copernicus global land service to the cloud~\citep{abernathey2021opening}

@@

Mature zarr implementations: Zarr-Python~\citep{zarrpython},
TensorStore~\citep{tensorstore}

and more experimental~\citep[e.g.][]{xtensor_zarr}

@@

CMIP6 dataset in Google Cloud~\cite{cmip6_dataset}
